--- 
title: "Exam project"
author: "Marco Zullich"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
documentclass: book
---

# Prerequisites

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(rstan)
library(bayesplot)

data = mlmRev::ScotsSec
```

# Problem specification

```
The R package mlmRev contains the ScotsSec dataset on scores attained by Scottish
secondary school students on a standardized test taken at age 16.
The data include 3,435 observations on 6 variables. The help file
description is as follows:

    * verbal The verbal reasoning score on a test taken by the students on entry
      to secondary school
    * attain The score attained on the standardized test taken at age 16
    * primary A factor indicating the primary school that the student attended
    * sex A factor with levels M and F
    * social The student’s social class on a numeric scale from low to high social
      class
    * second A factor indicating the secondary school that the student attended

After performing some explorative analyses:

    1. Consider the binary variable attain01 which takes values 1 if attain is greater
       than 5 and 0 otherwise. Build a model for studying the effects of covariates on
       attain01 with rstan, taking into account the hierarchical structure of the data.
    2. Check the model fit and comment the results.
    3. Draw inference on school random effects. Does the primary school matter?
    4. [optional] Propose an alternative model for the variable attain
       (stan fit is not required).

```

# Preliminary analysis

## Univariate analysis

As usually done on preliminary analysis, the first thing we want to do is establishing the types of the variables: 

```{r}
inspectdf::inspect_types(data)
```

We have 3 categorial and 3 numerical variables.

### Numerical variables

First we'll analyze the numerical ones:
```{r}
data_inspect_num = inspectdf::inspect_num(data)
data_inspect_num %>% select(-hist)
```

```{r}
data_inspect_num %>% inspectdf::show_plot()
```

We see that the only variable which seems to have an approximately continuous distribution is `verbal`.

The variable `attain`, which will be our response, show a very imbalanced bimodal discrete distribution, with peaks around 2 and 10.

`social`, instead, despite seemingly having a large support, has appearantly only have few values:
```{r}
data$social %>% table
```

### Categorical variables

Next, we analyze the factors:
```{r}
inspectdf::inspect_cat(data) %>% inspectdf::show_plot()
```
From the plot above we can see that:

* The `primary` variable has a huge support (148 categories) and there seems not to be a prevalent school between them, although there're primary schools with very little representation.

    ```{r}
data %>% group_by(primary) %>% summarise(n=n()) %>% select(primary,n) %>%
    mutate(size=ifelse(n<10,"small",ifelse(n>50,"large","medium"))) %>%
    pull(size) %>% table
    ```
  As we can see, 43 primary schools are represented by 10 students or less.

* The `second` variable seems to have a very uniform distribution as well: we have 19 secondary schools in which the test is submitted to students
    ```{r}
    data %>% group_by(second) %>% summarise(n=n()) %>%
      filter(n==max(n) | n==min(n))
    ```
  The school 10 (smallest one) has 92 students, while the school 14 (biggest one) has 290.

* The students are equally distributed between females and males

Finally, as per request point 1, we have to recode the variable `attain` into `attain01` which is an indicator for the score test being higher than 5:
```{r}
data = data %>%
  mutate(attain01=cut(attain,breaks=c(-Inf,5,Inf),labels=c(0,1)))

ggplot(data, aes(x=attain01)) + geom_bar(fill='orange')
```

We see that the two classes are balanced (around 1750 students in class 0, 1600 in class 1).

### Missing values

We can easily test for missing values with `inspect_na` of the `inspectdf` package.

```{r}
inspectdf::inspect_na(data)
```

We have no missing values.

## Bivariate analysis

Before starting on with the bayesian modelling, it's a nice practice to check for correlation between the variables in the dataset.

### `attain` and `verbal` {#firstbiv}

The first correlation I'll check is `attain01` vs. `verbal`. Since the former is a dichotomic variable, and the latter is approximately continuous, I'll fit a frequentist logistic regression to check:

```{r firstglm}
glm(attain01~verbal, data=data, family=binomial) %>% summary
```

The coefficient for `verbal` is positive and significative, yet not huge: that suggests a small positive correlation between the score obtained during the secondary school's entrance test and the test whose scores are being analyzed now. This can be appreciated in the plots below:

```{r box_attain_verb}
ggplot(data=data, aes(y=verbal, x=attain01, fill=attain01)) +
  geom_boxplot() +
  ggtitle("Distribution of `verbal` given `attain01`")
```

Especially from the boxplot above, we can see that the distributions of `verbal` given `attain01` are different, and that the central 50% of the distribution lie on separate portions of the space of `verbal`.

Those observations might lay the foundations for the models we're going to construct in the next sections.

```{r, echo = F}
plot(data$attain01 %>% as.character %>% as.numeric ~ data$verbal,
     xlab = "verbal",
     ylab = "attain01",
     pch = 18,
     yaxt = "n",
     col = "darkgray")
curve(exp(.2+.16*x)/(1+exp(.2+.16*x)), add=T, col="red")

```


### `attain` and `sex`

First I create a contingency table, from which we can appreciate a difference in distribution of `attain01` for different sex groups:

```{r, warning=FALSE}
contmatr = data %>%
  group_by(sex, attain01) %>%
   summarise(n = n()) %>%
   spread(attain01, n) %>% 
   magrittr::set_rownames(.$sex) %>%
   ungroup() %>%
   select(-sex) %>%
   as.matrix

make_cont_table = function(x){
  rbind(
  cbind(x,apply(x,MARGIN = 1, FUN = sum) %>%
                  matrix(ncol = 1)),
  c(apply(x,MARGIN = 2, FUN = sum),
    1))
}

make_cont_table(contmatr / sum(contmatr))
                  
```

These two plots help better visualize the disparity:

```{r mosaic_attain_sex}
vcd::mosaic(contmatr, shade=T, legend=F,
            main = 'Joint distribution of `attain01` and `sex`',
            labeling_args = 
              list(set_varnames = c(A="sex", B="attain01")))

```

```{r}
ggplot(data, aes(x=attain01, fill=sex)) + 
  geom_bar() +
  ggtitle("Distribuion of `attain01` given `sex`")
```

From the charts above, there seems to be a disparity in results depending upon the gender: females seem to have fared better than men, despite the marginal distribution of `sex` being almost uniform 


### `attain` and social category

First, I recode the variable `social` into a factor with three determinations:

```{r}
data = data %>% mutate(social_cat = cut(
          social, c(-Inf,2,21,Inf), labels=c('low','middle','high')
        ))
```

Then, I analize it in similar fashion as the previous paragraph:


```{r, warning=FALSE}
contmatr = data %>%
  group_by(social_cat, attain01) %>%
   summarise(n = n()) %>%
   spread(attain01, n) %>% 
   magrittr::set_rownames(.$social_cat) %>%
   ungroup() %>%
   select(-social_cat) %>%
   as.matrix


make_cont_table(contmatr/sum(contmatr))
```

Charts:

```{r}
vcd::mosaic(contmatr, shade=T, legend=F,
            main = 'Joint distribution of `attain01` and social category',
            labeling_args = 
              list(set_varnames = c(A="social category", B="attain01")))

```

It seems that:

* there's a large disparity between results between low and medium/high class students: the former performed poorly (around 60% got a mark below 5), while the latter performed better (35% of middle/high class got a mark below 5).
* however, there seems to be a small difference between middle and high class students, since the middle class overall performed better than high class ones, at least in getting at least a sufficient mark in the test.

#### Other considerations

Adding other variables to the boxplot in \@ref(box_attain_verb), we see that:

* `sex` doesn't change much the distribution:

    ```{r}
    ggplot(data=data, aes(y=verbal, x=attain01, fill=sex)) +
      geom_boxplot() +
      ggtitle("Distribution of `verbal` given `attain01` and `sex`")
    ```

    That probably indicates that the correlation noted in \@(mosaic_attain_sex) may be (almost) fully explained by `verbal`, and the addition of sex to a model already containing `verbal` as predictor might not bring significative results.
    
    ```{r}
    glm(attain01~sex+verbal, data=data, family=binomial) %>%
      summary()
    ```
    
    There seems not to be a substantial difference in effect on `attain01` for males or females.

* adding the social category may be useful:

    ```{r}
    ggplot(data=data, aes(y=verbal, x=attain01, fill=social_cat)) +
          geom_boxplot() +
          ggtitle("Distribution of `verbal` given `attain01` and social category") +
      scale_fill_brewer(palette=5)
    
    ```

    ```{r}
    glm(attain01~social_cat+verbal, data=data, family=binomial) %>%
      summary()
    ```
    
    It seems that the social category adds something to the relationship between `atttain01` and `verbal`.

### `attain` and secondary school

```{r}
ggplot(data %>% select(attain01, second) %>% group_by_all %>%
         summarise(n = n()) %>%
         inner_join(data %>% select(second) %>%
                      group_by(second) %>%
                      summarise(stud = n())) %>%
         mutate(prop = n/stud),
       aes(x=second, fill=attain01)) + 
  geom_bar(aes(y=prop), position="stack", stat="identity") +
  ggtitle("Proportion of `attain01` for each secondary school")
```

There proportions of 0s and 1s in test sufficiency are etherogeneous across the schools: in school 19 only ~20% of students got more than 5 as a score, while in school 12 ~60% of student achieved this result. 

### social composition for primary schools

```{r}
ggplot(data %>%
         select(second, social_cat) %>% group_by_all %>%
         summarise(n = n()) %>%
         inner_join(data %>% select(second) %>%
                      group_by_all %>%
                      summarise(stud = n())) %>%
         mutate(prop = n/stud),
       aes(x=second)) + 
  geom_bar(aes(y=prop, fill=social_cat), position="stack", stat="identity") +
  ggtitle("Proportion of students from social classes for each secondary school") 
  
  
```

```{r}
data %>% filter(second == 19) %>% group_by(social_cat) %>% summarise(n=n())
```


### attain and primary school

```{r}

ggplot(data %>% select(primary, attain01) %>% group_by_all %>%
         summarise(n = n()) %>%
         inner_join(data %>% select(primary) %>%
                      group_by_all %>%
                      summarise(stud = n())) %>%
         mutate(prop = n/stud) ,
       aes(x=primary, fill=attain01)) + 
  geom_bar(aes(y=prop), position="stack", stat="identity") +
  ggtitle("Proportion of test results for each primary school") +
  theme(axis.text.x = element_blank()) 
```

# Bayesian modelling

## `attain01` ~ `verbal`

As first mentioned before, the first model I'll try to fit will be `attain01` ~ `verbal`.

I have already proven, in a frequentist framework, in \@ref(firstbiv), that there exists a correlation between the two variables.
This will be tested in a bayesian framework.

`attain01` is a dichotomic variable, hence the model that I'll fit is a logistic regression (with canonical link). I'll also pick a vague prior on the coefficients: being the sample size large, in all chances the likelihood will largely dominate on the prior.

This is the model specification:

$$
\text{attain01}_i \sim Bernoulli(\text{logit}^{-1}(\eta_i)) \\
\eta_i = \alpha + \beta \cdot \text{verbal}_i \\
\alpha \sim N(0,10) \\
\beta \sim N(0,2.5)
$$

Note that the priors for $\alpha, \beta$ have been chosen as the default for `stan_glm` of `rstanarm` package.

Model compile:
```{r, cache=T}
model_lin_1 = stan_model('models/stan_model_glm_1.stan')
```
 
and fit:
```{r, eval=F}
fit_lin_1 = sampling(model_lin_1,
                     data = list(N = data %>% nrow,
                                 x = data$verbal,
                                 y = data$attain01 %>%
                                   as.character %>%
                                   as.numeric),
                     cores = 4)

saveRDS(fit_lin_1, "models/fit_lin_1.rds")
```

```{r, echo=F, cache=T, results='hide'}
fit_lin_1 = readRDS("models/fit_lin_1.rds")
```



### Coefficients analyisis

These are the coefficients estimated by the model:
```{r}
print(fit_lin_1, pars = c("alpha","beta"))
```

As expected, the coefficients have the same mean as the frequentist glm's estimate (see \@ref(firstbiv)). Both have 0 well outside their 95% C.I.

The interpretation of $\beta$ does not change: it has a positive effect on determining the sufficiency in the test; for a unitary increase in `verbal`, there's an increase of .16 in the linear predictor, or in the log-odds of the probability of success.

### Posterior predictive checks

First, I'm going to have a look at the fit of the mass function of the replicated data to the original y:

```{r}
y = data$attain01 %>% as.character %>% as.numeric
y_rep = as.matrix(fit_lin_1, pars = "y_rep")

ppc_ecdf_overlay(y,y_rep,discrete = T)
```

A check can be done on the bernoulli's parameter $\theta$, which in our model is, on average $\text{logit}^{-1}(\alpha+x*\beta)$.

Being our observed and replicated data just vectors of 0s and 1s, the parameter in an observed or replicated array is the mean of the values of such array.

```{r}
ppc_stat(y,y_rep)
```

The fit is excellent also considering the parameter.

To have an estimate for residuals, I'll calcuate the Brier Score on the model:

```{r}
hist_fun = function(y, y_rep, fun){
  
  
  vals = y_rep %>% apply(1,function(x) fun(y, x) )
  
  print(paste0("average statistic = ",mean(vals)))
  print(paste0("sd = ",sd(vals)))
  
  return (ggplot(data = vals %>% as_tibble, aes(x=value)) +
    geom_histogram(aes(y=..density..),  fill = "orange") +
    geom_density() +
    xlab ( as.character(substitute(fun)) ) )
}

(brier_1 = hist_fun(y, y_rep, DescTools::BrierScore))
```

An average value of 0.28 indicates a moderate fit.


```{r}
(loo_1 = loo::loo(loo::extract_log_lik(fit_lin_1)))
```

## Hierarchical models

I provide two generic `stan` models for the following sections:

```{r}
model_lin_hier_interc = stan_model('models/stan_model_glm_2.stan', auto_write = T)

model_lin_hier_both = stan_model('models/stan_model_glm_3.stan', auto_write = T)

model_lin_hier_interc_s = stan_model("models/stan_model_glm_2_pri_strong.stan")
```

### Hierarchy over social category

```{r, eval=F}
fit_lin_hier_soc_interc = sampling(model_lin_hier_interc, 
  data = list(N = data %>% nrow,
               x = data$verbal,
               y = y,
               G = data$social_cat%>%levels%>%length,
               group_mapping = data %>% pull(social_cat) %>%
                                as.numeric),
  cores = 4)

saveRDS(fit_lin_hier_soc_interc, "models/fit_lin_soc.rds")
```

```{r message=FALSE, warning=FALSE, include=FALSE, cache=T}
fit_lin_hier_soc_interc = readRDS("models/fit_lin_soc.rds")
```


```{r}
mcmc_intervals(as.matrix(fit_lin_hier_soc_interc), pars = 
                 dimnames(fit_lin_hier_soc_interc)$parameters %>%
                 as_tibble %>%
                 filter(grepl("alpha|beta",value)) %>% pull)
```

As expected, there're different values of $\alpha$ for medium and high class. $\beta$ is still significative with a very precise posterior.

```{r}
ppc_stat_grouped(y, yrep = as.matrix(fit_lin_hier_soc_interc,
                                    pars = "y_rep"),
                 group = data$social_cat)
```

```{r}
ppc_stat_grouped(y, yrep = as.matrix(fit_lin_hier_soc_interc,
                                    pars = "y_rep"),
                 group = data$social_cat, stat = sd)
```


```{r}
hist_fun(y, as.matrix(fit_lin_hier_soc_interc,
                                    pars = "y_rep"),
         DescTools::BrierScore)
```




```{r}
(loo_soc=loo::loo(loo::extract_log_lik(fit_lin_hier_soc_interc)))
```


#### Model with variable intercept and slope

To improve the result of the hierarchical model, I'll be fitting a model with variable intercept and slope over the same hierachy.


```{r linHierSexSoc, eval=F}
fit_lin_hier_soc_both = sampling(model_lin_hier_both, 
  data = list(N = data %>% nrow,
               x = data$verbal,
               y = y,
               G = data$social_cat%>%levels%>%length,
               group_mapping = data %>% pull(social_cat) %>%
                                as.numeric),
  cores = 4)

saveRDS(fit_lin_hier_soc_both, "models/fit_lin_4.rds")
```

```{r message=FALSE, warning=FALSE, include=FALSE, cache=T}
fit_lin_hier_soc_both = readRDS("models/fit_lin_4.rds")
```

```{r}
mcmc_intervals(as.matrix(fit_lin_hier_soc_both), pars = 
                 dimnames(fit_lin_hier_soc_both)$parameters %>%
                 as_tibble %>%
                 filter(grepl("alpha|beta",value)) %>% pull)
```

The model doesn't seem to add up to what we've seen before, because the $\beta$s have all a very similar posterior distribution. I'll have a look at the Brier score as a last check:

```{r}
(loo_soc_both=loo::loo(loo::extract_log_lik(fit_lin_hier_soc_both)))
```

### Hierarchy over primary school

#### Model with variable intercept, fixed slope

Having seen that the previous hierarchy did not offer much contribution to the model, I decided to change the hierarchical structure, focusing on the primary school of origin.

This model will begin in trying to answer to the request nr. 3:
```
Draw inference on school random effects. Does the primary school matter?
```

The model will be a hierarchical bayesian logistic regression with a random effect for the primary school:

```{r, eval=F}
fit_lin_hier_pri_interc = sampling(model_lin_hier_interc, 
  data = list(N = data %>% nrow,
               x = data$verbal,
               y = y,
               G = data$primary%>%levels%>%length,
               group_mapping = data %>% pull(primary) %>%
                                as.numeric),
  cores = 4)

saveRDS(fit_lin_hier_pri_interc, "models/fit_lin_2.rds")
```

```{r message=FALSE, warning=FALSE, include=FALSE, cache=T}
fit_lin_hier_pri_interc = readRDS("models/fit_lin_2.rds")
```

Let's analyze the posterior in an automated fashion:

```{r}
fit_lin_sum = summary(fit_lin_hier_pri_interc)
fit_lin_sum = fit_lin_sum$summary %>%
  as_tibble %>%
  tibble::add_column(param = fit_lin_sum$summary %>% row.names) %>% select(param, everything())
```

* $\beta$

    ```{r}
    fit_lin_sum %>% filter(param == 'beta')
    ``` 

    The mean is very close to the non-hierarchical model ($\Delta \pm 0.02$), and 0 is well outside the 95% CI.

* Intercepts

    Instead of printing directly the intercepts' posteriors, I'll store them in a data frame and make considerations on the 95% CI via `dplyr`.
    
    First, I create the dataframe and add a new variable telling me whether 0 is or is not within the 95% CI of each parameter:

    ```{r}
    fit_lin_sum_alphas = fit_lin_sum %>%
      filter(grepl('alpha',param)) %>% 
      mutate(sign = ifelse(`97.5%`*`2.5%`<0, "NoSign", "Sign"))
    ```

    Then, I group by the new variable to see how many school have a significative $\alpha$ parameter:
    
    ```{r}  
    fit_lin_sum_alphas %>% select(sign) %>% group_by(sign) %>%
      summarise(n=n())
    ```

    Out of 148 primary schools, only 25 schools show to have somewhat an effect on the score.
    
    These are such schools:

    ```{r}
    fit_lin_sum_alphas %>% filter(sign == 'Sign')
    ```

    ```{r}
    mcmc_intervals(as.array(fit_lin_hier_pri_interc), pars =
                     fit_lin_sum_alphas %>%
                     filter(sign == 'Sign') %>%
                     pull(param))
    ```

Analysis on some schools:
    
```{r}
data %>% filter(primary %in% c(3, 139, 69, 49)) %>%
  group_by(primary, attain01) %>% summarise(n=n())
```


The Brier score:

```{r}
hist_fun(data$attain01%>%as.character%>%as.numeric,
         as.matrix(fit_lin_hier_pri_interc, pars = "y_rep"),
         DescTools::BrierScore)
```

The model improves the Brier Score (ca 0.25 vs 0.28 for the non-hierarchical model): however, we can see that this improvement seems significative:

```{r, echo=F}
ggplot() +
  geom_histogram(data = y_rep %>% apply(1,function(x)
    DescTools::BrierScore(y, x)  ) %>% as_tibble ,
    aes(x=value,y=..density.., fill='blue') , alpha=.3) + 
  geom_histogram(data = as.matrix(fit_lin_hier_pri_interc, pars = "y_rep")  %>% apply(1,function(x)
    DescTools::BrierScore(data$attain01%>%as.character%>%as.numeric, x)  ) %>% as_tibble,
                 aes(x=value,y=..density.., fill='orange') , alpha=.3) + 
  scale_fill_discrete(name = 'Model',
                      labels = c("pooled", "hier. prim."))
```

This is not though confirmed by the LOOIC:

```{r}
(loo_pri=loo::loo(loo::extract_log_lik(fit_lin_hier_pri_interc)))
```

```{r}
loo_soc
```

```{r}
loo_1
```

Since the LOOIC operates on the predictive level, it's possible that a model showing increased goodness of fit but decreased LOOIC may be overfitting the data.

#### Stronger priors

Fit a new model with stronger priors:

$$
\alpha \sim N(0.2,1) \\
\beta \sim N(0,1)
$$

```{r, eval=F}
fit_lin_hier_pri_interc_s = sampling(model_lin_hier_interc_s, 
  data = list(N = data %>% nrow,
               x = data$verbal,
               y = y,
               G = data$primary%>%levels%>%length,
               group_mapping = data %>% pull(primary) %>%
                                as.numeric),
  cores = 4)

saveRDS(fit_lin_hier_pri_interc_s, "models/fit_lin_2s.rds")
```

```{r message=FALSE, warning=FALSE, include=FALSE, cache=T}
fit_lin_hier_pri_interc_s = readRDS("models/fit_lin_2s.rds")
```


* Analysis on $\alpha$s:

    ```{r, cache=T}
    fit_lin_sum = summary(fit_lin_hier_pri_interc_s)
    fit_lin_sum = fit_lin_sum$summary %>%
      as_tibble %>%
      tibble::add_column(param = fit_lin_sum$summary %>% row.names) %>% select(param, everything())
    ```

    ```{r}
    fit_lin_sum = fit_lin_sum %>% mutate(sig = ifelse(`2.5%`*`97.5%`<=0,0,1))
    
    print(fit_lin_sum %>% filter(grepl("alpha",param), sig==0) %>% count)
    print(fit_lin_sum %>% filter(grepl("alpha",param), sig==1) %>% count)
    ```

    ```{r}
    mcmc_intervals(as.array(fit_lin_hier_pri_interc_s), pars =
                         fit_lin_sum %>%
                         filter(sig == 1, grepl("alpha",param)) %>%
                         pull(param))
    ```



```{r}
loo_pri_s = loo::loo(loo::extract_log_lik(fit_lin_hier_pri_interc_s))
print(loo_pri)
print(loo_1)
print(loo_soc)
```


#### Model with variable intercept over primary schools grouped together in clusters

This model is done merely for showing how, by reducing the number of parameters, we can reduce overfitting in the model:

```{r}
data = data %>% mutate(school_clu = 
            ifelse(primary %in% c(16,69,139), 1,
            ifelse(primary %in% c(6,37,47,49,61,88,99,116,132,143), 2,
                   3)))

data$school_clu = data$school_clu %>% as.factor
```

```{r, eval=F}
fit_lin_hier_clu_interc = sampling(model_lin_hier_interc_s, 
  data = list(N = data %>% nrow,
               x = data$verbal,
               y = y,
               G = data$school_clu%>%levels%>%length,
               group_mapping = data %>% pull(school_clu) %>%
                                as.numeric),
  cores = 4)

saveRDS(fit_lin_hier_clu_interc, "models/fit_lin_clu.rds")
```


```{r message=FALSE, warning=FALSE, include=FALSE, cache=T}
fit_lin_hier_clu_interc = readRDS("models/fit_lin_clu.rds")
```

Let's draw the posteriors' graph like before:

```{r}
mcmc_intervals(fit_lin_hier_clu_interc %>% as.array, pars =
                 dimnames(fit_lin_hier_clu_interc)$parameters %>%
                 as_tibble %>%
                 filter(grepl("alpha|beta",value)) %>% pull)
```

The previously defined clusters behave as expected:

* the first group has a large positive coefficient with large variance
* the second group has a large negative coefficient with large variance
* the fifth group has a positive coefficient with mean very close to 0 (0.14), although 0 is not within the 95% CI (which starts at 0.04)

Moreover, $\beta$ has a value which is very close to that obtained in all of the previous models.

```{r}
ppc_stat_grouped(y, yrep = as.matrix(fit_lin_hier_clu_interc, pars = "y_rep"), group = data$school_clu )
```



```{r}
hist_fun(data$attain01%>%as.character%>%as.numeric,
         as.matrix(fit_lin_hier_clu_interc, pars = "y_rep"),
         DescTools::BrierScore)
```

This model shows a worse fit than the previous, but that might be attributed to the previous one overfitting the data.

```{r, cache=T}
loo::compare(
  loo::loo(loo::extract_log_lik(fit_lin_hier_pri_interc)),
  loo::loo(loo::extract_log_lik(fit_lin_hier_clu_interc))
)
```

The LOOIC large positive difference tells us that this model with clustered schools has a larger predictive power than the previous one.

### Hierarchy over secondary school

As a final model class, I'll try to fit a model with hierarchy over the secondary school instead of the primary.
My goal is to check whether we can use the primary school provenance or the secondary school belonging to predict better whether the test will be passed or not.


```{r, eval=F}
fit_lin_hier_sec_interc = sampling(model_lin_hier_interc, 
  data = list(N = data %>% nrow,
               x = data$verbal,
               y = y,
               G = data$second%>%levels%>%length,
               group_mapping = data %>% pull(second) %>%
                                as.numeric),
  cores = 4)

saveRDS(fit_lin_hier_sec_interc, "models/fit_lin_sec.rds")
```

```{r message=FALSE, warning=FALSE, include=FALSE, cache=T}
fit_lin_hier_sec_interc = readRDS("models/fit_lin_sec.rds")
```

```{r}
mcmc_intervals(fit_lin_hier_sec_interc %>% as.array, pars =
                 dimnames(fit_lin_hier_sec_interc)$parameters %>%
                 as_tibble %>%
                 filter(grepl("alpha|beta",value)) %>% pull)
```


```{r, cache=T}
loo_sec =  loo::loo(loo::extract_log_lik(fit_lin_hier_sec_interc))
print(loo_sec)
print(loo_1)
print(loo_soc)
```

### Cross-hierarchy with secondary school and social extraction

```{r}
data = data %>% mutate(soc_hi = factor(ifelse(social_cat=="low",0,1)))
data = data %>% mutate(soc_sec=factor(tidyr::unite(data,"soc_sec",c('soc_hi','second'))$soc_sec))

data %>% pull(soc_sec) %>% table %>% as.matrix
```

```{r, eval=F}
fit_lin_hier_soc_sec_interc = sampling(model_lin_hier_interc_s, 
  data = list(N = data %>% nrow,
               x = data$verbal,
               y = y,
               G = data$soc_sec%>%levels%>%length,
               group_mapping = data %>% pull(soc_sec) %>%
                                as.numeric),
  cores = 4)

saveRDS(fit_lin_hier_soc_sec_interc, "models/fit_lin_soc_sec.rds")
```

```{r message=FALSE, warning=FALSE, include=FALSE, cache=T}
fit_lin_hier_soc_sec_interc = readRDS("models/fit_lin_soc_sec.rds")
```

```{r, fig.height=10}
mcmc_intervals(fit_lin_hier_soc_sec_interc %>% as.array, pars =
                 dimnames(fit_lin_hier_soc_sec_interc)$parameters %>%
                 as_tibble %>%
                 filter(grepl("alpha|beta",value)) %>% pull)
```


```{r, cache=T}
loo_soc_sec = loo::loo(loo::extract_log_lik(fit_lin_hier_soc_sec_interc))
print(loo_soc_sec)
print(loo_1)
print(loo_soc)
print(loo_sec)
```


# Conclusion

During this project, the dataset ScotsSec was analyzed. It contains the marks (1-10) obtained by a group of Scottish secondary school students, coming from a set of 19 schools, at a test.

The following analyses were performed:

1. Relationships between the main variables in this model
  * I recoded variable `attain` from a discrete one with 10 terminations to a dichotomic one, which can be interpreted as passed/not passed the exam
  * There seems to be a clear correlation between `attain01` and `verbal`, the result of a test conducted on the students before accessing the secondary school. The correlations between `attain01` and `sex` seems spurious, almost fully explained by `verbal`; the `class` variable, instead, if recoded into a factor, shows some correlation with `attain01` even when `verbal` is taken into account.
2. Fit of a bayesian logistic linear regression for `attain01`:
  * with only `verbal` as a covariate (called **base model**): shows a nice fit, posterior predictive checks are good
  * adding a hierarchy over social class: improves model predictive performance
  * adding a hierarchy over primary school provenence (with fixed slope): shows improved fit, but predictive check using LOOIC isn't consistent with this; 
    + using informative prior reduces the number of "significative" coefficients, but LOOIC improves by a small margin over **base model**; still can't beat model with `verbal` and social extraction;
  * adding a hierarchy over secondary school with informative prior: intercepts are very small and most of them not significative, doesn't improve LOOIC w.r.t. the **base model**

The best model considered the observations above, is the following:

$$
i\in \{1,...,3435\} \\
\text{attain01}_{i} \sim Bernoulli(\text{logit}^{-1}(\eta_i)) \\
\eta_i = \alpha_{g(i)} + \beta \cdot \text{verbal}_i \\
\alpha \sim N(0,10) \\
\beta \sim N(0,2.5) \\
g(i) \text{ maps unit } i \text{ to its class extraction } j,\ \ j\in\{1,2,3\}
$$

```{r message=FALSE, warning=FALSE, echo=FALSE}
y_rep = as.matrix(fit_lin_hier_soc_interc,
                                    pars = "y_rep")[1:200,]


vals = y_rep %>% apply(1,function(x)pROC::auc(pROC::roc(y, x)))
  
print(paste0("average statistic = ",mean(vals)))
print(paste0("sd = ",sd(vals)))
  
ggplot(data = vals %>% as_tibble, aes(x=value)) +
    geom_histogram(aes(y=..density..),  fill = "blue") +
    geom_density() +
    xlab ( "AUC" )
```

## Possible improvements

Possible improvements in analyzing this dataset:

* Fit a logistic regression with Probit link instead of logit$^{-1}$. The generic `stan` file for it is included in the appendix.

* Consider the variable `attain`. Its distribution is:
  + discrete with values from 1 to 10
  + multimodal with spikes at 2 and 10
  
  To model directly this as a likelihood, two following approaches can be considered:
  + Model `attain` as a multinomial random variable with 10 possible outcomes:
  
  $$
  \mathbb{p} = [p_1, ..., p_{10}]^T,\ \sum_{j=1}^{10}p_j=1 \\
  y \sim Multi(n, \mathbb{p}) 
  $$
    One of the outcomes (e.g. 1) is elected as a baseline, and each of the other outcomes have their own linear predictor:
    
  $$
  \eta_j = X\beta^{(j)}
  $$
    Then
    
  $$
  \mathrm{B} = \{\beta^{(j)}\}_{j\in \{1,...,10\}} \\
  \beta^{(1)}=0 \text{ baseline} \\
  p(y|\mathrm{B}) = \prod_{j=1}^{10}\frac{\exp(\eta_j)}{\sum_{k=1}^{10} \exp(\eta_k)}
  $$
  
  + Model `attain` as a mixture of densities:
  
  1. Transform this variable as $z_i = \frac{\text{attain}_i - 1}{9}$
  2. Suppose it continuous
  3. Take its likelihood to be a mixture of $K$ betas, each with its own $\alpha_k, \beta_k$ parameters:
  
  $$
  \rho_1,...,\rho_k: (\rho_i\geq 0) \wedge (\sum_k \rho_k = 1) \\
  p(y_i|\alpha,\beta,\rho) = \sum_{k=1}^K \rho_k\cdot 
                                  p_k(y_i|\alpha_k,\beta_k)
  $$
  
  4. Fix priors on $\alpha, \beta, \rho$.
  
  
# Appendix - Stan files

## Simple logistic regression with one predictor

```{r}
cat(readLines('models/stan_model_glm_1.stan'), sep = '\n')
```

## Hierarchical logistic regression with one predictor and group, variable intercept

```{r}
cat(readLines('models/stan_model_glm_2.stan'), sep = '\n')
```

### More informative priors

```{r}
cat(readLines('models/stan_model_glm_2_pri_strong.stan'), sep = '\n')
```

## Hierarchical logistic regression with one predictor and group, variable intercepts and slopes

```{r}
cat(readLines('models/stan_model_glm_3.stan'), sep = '\n')
```

## Simple probit regression with one predictor

```{r}
cat(readLines('models/stan_model_glm_1probit.stan'), sep = '\n')
```

  