### Cross-hierarchy over sex and social category

#### Model with variable intercept, fixed slope

First, I need to add the column corresponding to the cross-hierarchy in my data:
  
  ```{r}
data = data %>% tibble::add_column( hier =
                                      tidyr::unite(data,"hier",c('sex','social_cat'))$hier
)
data$hier = data$hier %>% as.factor
```

Then, I can fit the model

```{r, eval=F}
fit_lin_hier_sex_soc_interc = sampling(model_lin_hier_interc, 
                                       data = list(N = data %>% nrow,
                                                   x = data$verbal,
                                                   y = y,
                                                   G = data$hier%>%levels%>%length,
                                                   group_mapping = data %>% pull(hier) %>%
                                                     as.numeric),
                                       cores = 4)

saveRDS(fit_lin_hier_sex_soc_interc, "models/fit_lin_3.rds")
```

```{r message=FALSE, warning=FALSE, include=FALSE, cache=T}
fit_lin_hier_sex_soc_interc = readRDS("models/fit_lin_3.rds")
```


```{r}
mcmc_intervals(as.matrix(fit_lin_hier_sex_soc_interc), pars = 
                 dimnames(fit_lin_hier_sex_soc_interc)$parameters %>%
                 as_tibble %>%
                 filter(grepl("alpha|beta",value)) %>% pull)
```

The intercepts seem to be significative only for class 3 and 6 (corresponding to low social category for both males and females)

```{r}
hist_fun(data$attain01%>%as.character%>%as.numeric,
         as.matrix(fit_lin_hier_sex_soc_interc, pars = "y_rep"),
         DescTools::BrierScore)
```

The Briar score is improved by too little over the non-hierarchical mode.

```{r}
loo::compare(
  loo_1,
  loo::loo(loo::extract_log_lik(fit_lin_hier_sex_soc_interc))
)
```

The LOOIC instead tells us that there's a moderate improvement in predictive power by the latter model.
